Information carried by thalamocortical axons enables self-organization of the barrel field

### 2019-12-02 paper/briefpaper/brief.tex
WIP reducing text down

\begin{abstract}
Understanding how genetic information and spontaneous pattern formation
together give rise to the whisker barrels is an important challenge for
developmental neurobiology. The barrel patterning that coalesces in layer IV
has been described formally as a Voronoi tesselation, which is indicative of
an underlying developmental process based on local competitive
interactions. If pattern formation relies only on such intrinsic
self-organising processes, how can the somatotopic order at the periphery be
maintained in the projections from brainstem to thalamus and from thalamus to
cortex?
%
To investigate, we began by extending a existing computational model of axon
branching and synaptogenesis in thalamocortical projections under the
influence of molecular signalling gradients.  We found that, although these
processes alone are incapable of generating patterns of cortical fields that
are more intricate than the patterns already present in the signalling
gradients, the addition of competition between the branches of axons that
originate from different thalamic barreloids allowed a full barrel-field
pattern to emerge in the presence of only two gradients.
%
%However, a
%full barrel-field pattern emerges under the guidance of as few as four
%signalling gradients when the model is extended to introduce competition
%between the branches of axons that originate from different thalamic
%barreloids.
%
%We show that the key requirements for the emergence of realistic
%(Voronoi tesselation) patterning are i) at each cortical location
%thalamocortical projections compete for a limited number of available
%connections, ii) at each location the axonal branching rate of each
%thalamocortical projection is reduced by the total branching density of the
%other projections, and iii) the density of axon branching for each
%thalamocortical projection is conserved.
%
We demonstrate how a purely
self-organising system can initialise a pattern which can then be faithfully
replicated in multiple neural structures.
%We show that this minimal model of
%barrel self-organisation provides a parsimonious account of data from a range
%of classic studies in which intrinsic and extrinsic factors have been
%experimentally manipulated.
\end{abstract}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the
constraints under which brains evolve and develop.
%Whether or not patterning
%governs information processing, models of the development of information
%processing circuits can be validated in terms of their ability to account for
%biological patterning \citep{purves_iterated_1992}.
%%see also \citealp{wilson_what_2015,bednar_cortical_2016}
One of the most
distinctive patterns of neural connectivity is found in the rodent
barrel cortex \cite{woolsey_structural_1970}.
%(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974},
%see \citealp{fox_barrel_2008}).
Understanding how this particular pattern forms might shed light on the
interplay of genetic and extrinsic factors that shape cortical circuits.
%, and on a wealth of
%anatomical, physiological, and behavioural data obtained using this
%model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
primary somatosensory cortex as dense clusters of thalamocortical axons, which
are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical
surface the barrels constitute a map of the arrangement of the whiskers on the
face,
%(see \citealp{yamakado_subdivision_1979})
such that the cells of adjacent barrels respond most strongly and most quickly
to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. The
barrels reflect subcortical whisker maps comprising cellular aggregates known
as barreloids in the thalamus \cite{van_der_loos_barreloids_1976} and
barrelettes in the brainstem.
%\citep{killackey_pattern_1980}.

%Each whisker and its corresponding barrelette, barreloid, and barrel can be
%identified by its position on a two-dimensional grid.
%\citep{zucker_coding_1969,killackey_pattern_1980,van_der_loos_barreloids_1976,haidarliu_size_2001}.
%
%The barrels are normally up to twice as long in the axis aligned to the grid
%‘rows’ compared to the ‘arcs’, and the rows are visible from day 2
%\citep{rebsam_refinement_2002} whereas the arcs become distinct by day 3
%\citep{erzurumlu_development_2012,rebsam_refinement_2002}.
%In metabotropic glutamate receptor 5 (mGlu5) knockouts the barrel rows develop
%but the arcs do not (\citealp{hannan_plc-1_2001} see also
%\citealp{fox_barrel_2008}), thus distinct processes likely guide development
%of the barrels with respect to the orthogonal grid axes.

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
\citep{anton-bolanos_prenatal_2019}.
%(\citealp{anton-bolanos_prenatal_2019}, see also
%\citealp{anton-bolanos_developmental_2018}).
Patterning also depends on a complex network of axon guidance processes, which
includes interactions between guidance molecules like ephrin-A5 and A7
\citep{miller_epha7-ephrin-a5_2006} and between adhesion molecules like
cadherin-6 and 8 \citep{bishop_regulation_2000}, and the
binding of p75 to neurotropins
\citep{bishop_distinct_2002}.
% or both citations?
%\citep{shimogori_fibroblast_2005,bishop_distinct_2002}.
%
%(\citealp{shimogori_fibroblast_2005,bishop_distinct_2002}
%; see also \citealp{dye_lifespan_2011,dye_lifespan_2011-1}).
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17
and transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate \cite{assimacopoulos_fibroblast_2012}.
% A longer list of citations:
%\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012,borello_sp8_2014,sahara_sp8_2007,ypsilanti_transcriptional_2016,sur_patterning_2005}.
%
While complex structure in protein expression can be observed for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008}, we are not aware of
any report of gene expression patterns with the complexity of the barrel field
and so the mechanism by which the barrels form seems likely to be one in which
the structure forms by self-organization.

The boundaries of the barrels form a Voronoi tesselation across the cortical
sheet \citep{senft_mouse_1991}. This patterning suggests that the barreloid
topology may be preserved in the initial projection of the thalamocortical
axons into the cortex, and that a barrel forms by the lateral branching of the
axons from their initial centres of mass, which ceases upon contact with the
axons branching from adjacent centres. According to this model the barrel
pattern could be entirely determined by the distribution of the initial centre
points, which would itself be determined by guidance cues. However, the
assumption of pre-arranged centres is difficult to resolve with the
observation that thalamocortical axons arrive in the cortical plate as an
undifferentiated bundle prior to barreloid formation within the thalamus
\cite{agmon_organized_1993}. % Plus erzurumli, probably

Alternatively, reaction-diffusion dynamics could generate a Voronoi-like
tesselation in the cortex without the need for pre-arranged centres, by
amplifying characteristic modes in the noisy initial distribution of bundled
axon branches, as a net effect of short-range cooperative and longer-range
competitive interactions between cells. According to this model the barrel
pattern would be determined by the relative strength of these interactions and
by the shape of the cortical field boundary. However, reaction-diffusion
dynamics alone are unlikely to be able to account for the precise
co-registration and topographic connectivity between emergent domains in
thalamus and cortex, the non-uniform sizes and spacing of the barrels to form
rows versus arcs, and the influence of gene expression gradients.

As theories of barrel patterning, the centre-point and reaction-diffusion
models are not mutually exclusive. Pre-organised centres in the initial
conditions could bias reaction-diffusion processes to generate the specific
somatotopic arrangement of barrels more reliably, and similarly the mechanisms
by which axons branch laterally from barrel centres may involve a tension
between cooperative and competitive interactions. However, a proof that barrel
patterning can in principle emerge within an undifferentiated bundle of
thalamocortical axons, based only on local interactions between cells, would
remove the need to invoke a global supervisor, i.e., to assume a separate
stage and/or extrinsic mechanism for pre-organising thalamocortical
connections.

We seek to establish whether it is possible for the observed arrangement of
barrels to self-organise in a system governed by reaction-diffusion dynamics,
under the guidance of molecular signalling gradients, and in the absence of
pre-defined centre-points.
%
Our approach is to represent in a computational model the minimal set of
biological assumptions that might lead to the emergence of barrel patterning
in simulation, and, if necessary, to add to this minimal set only assumptions
that can be represented by the local exchange of information between adjacent
cells. The emergence of barrel patterning under such conditions would serve as
an existence proof of the plausibility of the hypothesis that barrel
development is a self-organising process.

>>>>

\begin{abstract}
The whisker barrel patterning that coalesces in cortical layer IV has been
described formally as a Voronoi tesselation, which is indicative of an
underlying developmental process based on local competitive interactions. If
pattern formation relies only on such intrinsic self-organising processes, how
can the somatotopic order at the periphery be maintained in the projections
from brainstem to thalamus and from thalamus to cortex?
%
To investigate, we began by extending a existing computational model of axon
branching and synaptogenesis in thalamocortical projections under the
influence of molecular signalling gradients.  We found that, although these
processes alone are incapable of generating patterns of cortical fields that
are more intricate than the patterns already present in the signalling
gradients, the addition of competition between the branches of axons that
originate from different thalamic barreloids allowed a full barrel-field
pattern to emerge in the presence of only two gradients.
%

%However, a
%full barrel-field pattern emerges under the guidance of as few as four
%signalling gradients when the model is extended to introduce competition
%between the branches of axons that originate from different thalamic
%barreloids.
%
%We show that the key requirements for the emergence of realistic
%(Voronoi tesselation) patterning are i) at each cortical location
%thalamocortical projections compete for a limited number of available
%connections, ii) at each location the axonal branching rate of each
%thalamocortical projection is reduced by the total branching density of the
%other projections, and iii) the density of axon branching for each
%thalamocortical projection is conserved.
%
We demonstrate how a purely
self-organising system can initialise a pattern which can then be faithfully
replicated in multiple neural structures.
%We show that this minimal model of
%barrel self-organisation provides a parsimonious account of data from a range
%of classic studies in which intrinsic and extrinsic factors have been
%experimentally manipulated.
\end{abstract}

@@

\dropcap{O}ne of the most distinctive patterns of neural connectivity is found
in the rodent \emph{barrel cortex} \cite{woolsey_structural_1970}, which is
comprised of large cortical columns which develop from birth
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical
surface the barrels constitute a map of the arrangement of the whiskers on the
face, such that the cells of one barrel respond most strongly to the
deflection of its corresponding whisker \citep{armstrong-james_flow_1992}. The
barrel field reflects similar, subcortical whisker maps comprising cellular
aggregates known as \emph{barreloids} in the thalamus
\cite{van_der_loos_barreloids_1976} and \emph{barrelettes} in the brainstem.
%
In this paper, we consider the mechanism by which the map can be reliably
re-created in these separate tissues. Any candidate mechanism must maintain
the somatotopic ordering of the facial whiskers; it must operate in tissues
comprised of differing neuronal cell species and which develop at several
different length scales. We focus on the thalamic barreloid map and how
thalamocortical axons become ordered in their terminations in the barrel
field, but point out that the mechanism should be transferable to the creation
of the barreloids from the barrelettes and the formation of the barrelette
fields from the mystacial whiskers themselves.

The formation of the cortical barrels is related to the guidance of axons
which grow from the thalamus, via the internal capsule and into the developing
cortical sub-plate \citep{erzurumlu_development_2012}. Here, they form an
undifferentiated bundle \citep{agmon_organized_1993}, out of which they ascend
in an ordered pattern to arrive in cortical layer IV, where they form
synapses. Axon guidance depends on a network of actors including guidance
molecules such as ephrin-A5 and A7 and their receptors the Ephs
\citep{miller_epha7-ephrin-a5_2006} and adhesion molecules like cadherin-6 and
8 \citep{bishop_regulation_2000}. This network is orchestrated by interactions
between morphogens Fgf8 and Fgf17 and transcription factors Emx2, Pax6, Sp8,
and Coup-tf1 \citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which
are expressed in gradients that mark orthogonal axes and can be manipulated to
cause the barrels to stretch, shrink, shift, skew, and even duplicate
\cite{assimacopoulos_fibroblast_2012}.
%
While complex structure in protein expression can be observed for some genes,
such as the molecule Bhlhb5 \citep{joshi_bhlhb5_2008}, all of the factors
mentioned above are expressed in monotonic, slowly varying patterns. This
suggests that the barrel field is not created according to a genetically
specified `stencil' but is instead formed by a self-organising process.
%
% Can I get this to be less of an orphan sentence?
Barrel formation also requires afferent input, driven by whisker stimulation
and by thalamic calcium waves that propagate via mGlu5-receptor activation
\citep{anton-bolanos_prenatal_2019}.

The boundaries of the barrels form a Voronoi tesselation across the cortical
sheet \citep{senft_mouse_1991} in which each \emph{Voronoi domain} (a barrel)
contains a geometrically defined centre point. A Voronoi tesselation can be
formed by a process which is initialized with the locations of all the centre
points and then builds the Voronoi domains. The advantage of this, in
understanding barrel formation, is that by pre-specifying the centre-point
locations, the map is guaranteed to maintain its ordering. However, this
theory requires that some initial process defines the location of each barrel,
prior to the arrival of the thalamocortical axons. If this were so, then some
factor should be detectable which would indicate the structural complexity of
the barrel field.

The presence of the Voronoi tesselation could alternatively suggest a process
which forms the Voronoi domains according to local interactions between axon
branches, dendrites and the surrounding cells - such as in a reaction
diffusion system \cmnt{finish me}. For the barrels, this leaves
unanswered the question of how the somatotopy is preserved, as fluctuations in
such a process would likely cause a different pattern to form in each
individual.

We seek to establish whether it is possible, and if so, the minimal set of
biological assumptions required, for the observed arrangement of barrels to
self-organise in a system governed by reaction-diffusion dynamics, under the
guidance of molecular signalling gradients, and in the absence of pre-defined
centre-points.
